Uranyl solvation by a three-dimensional reference interaction site model

### 2015-04-07 doc/uo2_paper/uo2-II.tex
AM: re-apply figures from cwd
AM: regenerate tex fro JPCA

%% LyX 2.0.6 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,5p,sort&compress,times]{elsarticle}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{esint}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
%\usepackage{mathptmx}
%\usepackage{times}
%\def\thetable{\arabic{table}}
\usepackage{microtype}
\journal{Computational and Theoretical Chemistry}

@@

\usepackage{babel}
\begin{document}

\title{Uranyl solvation by the three dimensional reference interaction site
model}

\author[tum]{Alexei Matveev\corref{corr}}

\ead{alexei.matveev@gmail.com}

\author[tum]{Bo Li}

\ead{fixme@abc.com}

\author[tum,crc]{Notker Rösch}

\ead{fixme@xyz.com}

\cortext[corr]{Corresponding author}

\address[tum]{Department Chemie, Technische Universität München, 85747 Garching,
Germany}

\address[crc]{Catalysis Research Center, Technische Universität München, 85747
Garching, Germany}
\begin{abstract}
We report an implementation of the three-dimensional reference interaction
site model (3D RISM) specifically describing the treatment of the
long-range Coulomb field of charged species represented by point charges
and/or distributed charge density. A comparison of 1D and 3D results
for atomic ions demonstrates the reasonable accuracy even for moderate
box size and grid resolution. In an application to uranyl aqua complexes
with 4\textendash{}6 explicit aqua ligands and an implicit bulk solvent
modeled by RISM, we demonstrate that the 3D technique is not susceptible
to the deficiencies of 1D technique exposed in our previous work {[}Li,
Matveev, Krüger, Rösch, \emph{Comp. Theor. Chem.} \textbf{1051}, p.~151,
(2015){]}. The 3D method eliminates the artificial superposition of
explicit aqua ligands and the RISM medium. It predicts essentially
the same values for uranyl and uranyl-water bond lengths as a state
of the art polarizable continuum model. We find that with the first
solvation shell treated explicitly, the observables are nearly independent
of the closure order. The water exchange activation barrier obtained
in a hybrid approach combining the quantum mechanical description
of the uranyl interaction within the first solvation shell and the
RISM model for the bulk solvent is in close agreement with the experimental
and the best theoretical estimates. \end{abstract}
\begin{keyword}
Reference interaction site model, uranyl, aqueous solution
\end{keyword}
\maketitle

@@

Thermodynamic and structural properties of the uranyl(VI) cation,
UO$_{2}^{2+}$, the most common uranium species in aqueous solution
are crucial for understanding its environmental chemistry as well
as chemical processes in radioactive waste disposal and recycling
of used nuclear fuel \citep{book:Morss06,book:Grenthe92,Silva95}.
Quantum mechanics (QM) and classical molecular mechanics (MM) simulations
provide the two complimentary ways to address microscopic structure
and energetics of solvated uranyl species. Till now, the most popular
strategy to combine the accuracy of QM methods and the complexity
of the statistical solvent description is to apply a polarizable continuum
model (PCM) to represent the effective electrostatic field of the
solvent, usually employed in combination with phenomenological corrections
for dispersion-repulsion interactions and cavity formation energy
using empiric parameters \citep{Tomasi05,Fuchs02}. The state of the
art PCM is owing its success in part to the conductor-like screening
model (COSMO) of electrostatic interactions \citep{Klamt95} and has
successfully been applied in various studies of uranyl complexes \citep{Fuchs02,Moskaleva04,Cao05,Shamov05,Gutowski06}.
On the other hand, there is a significant effort to improve the empirical
model potentials for MM simulations, such as molecular dynamics (MD)
or Monte-Carlo (MC) methods, that yield the most explicit statistical
solvent description \citep{Guilbaud93,Guilbaud96,Rai12,Pomogaev13,Kerisit13,Tiwari14}.
Though several studies \citep{Gordon00,hagberg05,Frick09,Tirler13}
attempt to combine QM and stochastic MM models such combinations will
inevitably lack the precise reproducibility  and efficiency of analytic
continuum models due to statistical sampling techniques. Yet more
demanding QM dynamic simulations \citep{Buehl05,Buehl06,Buehl11,Nichols08}
are restricted to comparably small numbers of solvent molecules.

Meanwhile, the integral equation theories of molecular liquids \citep{book:Hirata03,book:Hansen06},
may provide sufficient accuracy at an affordable cost for simulating
chemical processes in solution by employing standard numerical methods
on contemporary hardware. Two approaches of this type are the reference
interaction site model (RISM) \citep{Chandler72,Hirata81,Hirata82,Kinoshita97}
and its three-dimensional generalization (3D RISM) \citep{Beglov95,Kovalenko98}
which provide the solvent structure in form of site-site radial distribution
functions (RDFs) or three-dimensional distribution functions of solvent
sites in an external (solute) field, respectively. Both 1D and 3D
theories provide variational expressions for the solvation energy
in the form of the excess chemical potential for several closure types
\citep{Singer85,Kast08} and, hence, intuitive expressions for the
first derivatives. The dielectrically consistent RISM (DRISM) \citep{Perkyns92}
was designed to compensate the error of the gas-like dielectric response
of RISM for electrolytes and ionic solutes \citep{Perkyns94,Chiodo07,Chuev08,Joung13}.

A correct handling of the electrostatic field of a charge distribution
requires additional care when periodic boundary conditions are employed
to simulate finite systems \citep{Makov95}. It was understood early
that the long-range asymptotes of the radial correlation functions
induced by Coulomb interactions in RISM need a separate treatment
as well \citep{Ng74}. For 3D RISM the few practical approaches include
the background charge correction technique \citep{Kovalenko00b},
monopole correction \citep{Kloss08b}, and an exact re-summation for
a superposition of gaussian charges \citep{Perkyns10}. In this work
we present a generalization of the exact re-summation technique for
arbitrary solute charge densities.

The integral equation theory has been combined with Hartree--Fock
electronic structure theory for self-consistent field calculations
(RISM SCF) in the pioneering work of Ten-no, Hirata, and Kato and
has been first applied to formaldehyde and carbonyl compounds in water
\citep{Tenno93,Tenno94}. An extension to 3D RISM SCF was developed
for density functional theory (DFT) \citep{Kovalenko99} and \emph{ab
initio} calculations \citep{Kloss08}. The analytical gradients for
the self-consistent combination of 3D RISM and DFT \citep{Gusarov06}
simplify explorations of free energy landscapes. Some of the recent
developments of hybrid approaches combining integral equation theories
and quantum chemistry are discussed in the recent review by Sato \citep{Sato13}.

In this work, we first describe the implementation of the 3D RISM
approach and a numerical treatment of the long-range Coulomb field
of the charged solute, Section~\ref{sec:methodology}, and then validate
the implementation by comparing the results of 1D and 3D techniques
for alkali and halide ions, Section~\ref{sec:simple-ions}. We further
evaluate the performance of the method for the uranyl(VI) dication
in aqueous solution represented by aqua complexes $[\text{UO}_{2}(\text{H}_{2}\text{O})_{n}]^{2+}$,
($n=4-6$), using a force field, Section~\ref{sub:MM+RISM}, and
a quantum mechanical representation of interactions within the first
solvation shell, Section~\ref{sub:QM+RISM}. We also compare the
results to those of PCM calculations. The activation barrier of water
exchange in the first solvation shell is analyzed in Section~\ref{sub:water-exchange}
using QM and MM models of interactions in the first solvation shell.

@@

In the 3D RISM approach, the solvent sites are exposed to an external
field of a solute of general shape. The solute-solvent interactions
are thus assumed to be pairwise, which in this formulation excludes
\emph{polarizable} solutes. Further, the interactions are usually
approximated by superposition of pairwise site-site contributions.
The 3D RISM equations operating with site-specific distributions can
be written in the following form \citep{book:Hirata03}:

\begin{equation}
\begin{cases}
h=\chi\ast c\\
h=c+t\\
h=f(-v+t)
\end{cases}\label{eq:rsim-h-c-t}
\end{equation}
Here $h$, $c$, and $t$ are the vectors of total-, direct-, and
indirect correlation functions $h_{\alpha}(\vec{r})$, $c_{\alpha}(\vec{r})$,
and $t_{\alpha}(\vec{r})$. All are dimensionless unknowns of the
equations that will be represented on a uniform grid of points. The
greek subscripts, such as $\alpha$, enumerate solvent interaction
sites.

The solvent susceptibility $\chi$ in the first of the three Eqs.~\ref{eq:rsim-h-c-t}
is represented by its (square symmetric) matrix $\chi_{\alpha\gamma}(r)$
derived from the radial site-site correlation functions of the pure
solvent in 1D RISM equations for a specific temperature and solvent
number density $\rho$. \citep{book:Hirata03}. The star denotes the
convolution integral and the sum over corresponding matrix indices.
For that reason it is convenient to represent the solvent susceptibility
by its (dimensionless) Fourier image $\tilde{\chi}_{\alpha\gamma}(k)$
\citep{Chandler72,book:Hirata03}. The solvent susceptibility as a
function of either $r$ or $k$ is spherically symmetric by construction.

The second of the Eqs.~\ref{eq:rsim-h-c-t} introduces the indirect
correlation $t$ that is conveniently used in the closure relation
--- the third and last of Eqs.~\ref{eq:rsim-h-c-t}. Closure is a
local relation between the values of site potential $v_{\alpha}(\vec{r})$
and, in this case, the indirect correlation $t_{\alpha}(\vec{r})$
at each point $\vec{r}$ and the value of the direct correlation $h_{\alpha}(\vec{r})$
at this point. It is the only non-linear relation between the unknowns
$h$, $c$ and $t$ in Eqs.~\ref{eq:rsim-h-c-t}. The site-specific
potentials $v$ are measured in units of temperature and thus are
also dimensionless. For the purposes of this work we have chosen a
very specific, though popular, form of the closure which happens to
be a relation between $h$ and the sum $x=-v+t$ sometimes referred
to as a renormalized indirect correlation \citep{Kast08}:
\begin{equation}
f(x)=\begin{cases}
\mathrm{e}^{x}-1 & x\le0\\
\sum_{k=1}^{n}x^{k}/k! & x>0
\end{cases}\label{eq:closure-pse}
\end{equation}
In the limit of $n\to\infty$ one recovers the hypernetted chain (HNC)
closure. Another popular closure is obtained for $n=1$. It was proposed
by Kovalenko and Hirata (KH) for improving the numerical stability
of the equations \citep{Kovalenko99}. Higher order closures with
$n>1$ are referred to as partial series expansion closures of order
$n$ (PSE$n$). For consistency, we will also refer to the KH closure
as PSE1 in this work.

We used pair-wise interactions between solute and solvent sites $i$
and $\alpha$ to define the solute field: $v_{\alpha}(\vec{r})=\beta\sum_{i}u_{i\alpha}(|\vec{r}-\vec{r}_{i}|)$
where $\beta$ is the inverse temperature. Each potential $u_{i\alpha}$
is defined as the sum of the Coulomb interaction $u_{i\alpha}^{\text{C}}(r)=q_{i}q_{\alpha}a[f_{\text{CS}}(ar)+f_{\text{CL}}(ar)]$,
where we separated the singular short-range $f_{\text{CS}}(r)=\mathrm{erfc}(r)/r$
from the regular long-range $f_{\text{CL}}(r)=\mathrm{erf}(r)/r$
part, and a Lennard--Jones (LJ) type potential, $u_{i\alpha}^{\text{LJ}}(r)=\epsilon_{i\alpha}f_{\text{LJ}}(r/\sigma_{i\alpha})$
with a LJ form $f_{\text{LJ}}(r)=4(r^{-12}-r^{-6})$. Parameters of
the pair interactions were derived from site parameters using the
Lorentz--Berthelot rules \citep{book:Hirata03}. The Ewald range parameter
$a=1.2$~Å$^{-1}$ is common for all site pairs. Further, to circumvent
numerical problems with singularities the primitive forms $f_{\text{CS}}(r)$
and $f_{\text{LJ}}(r)$ were regularized by substituting them with
quadratic functions $\bar{f}(r)=f(b)+f^{\prime}(b)(r^{2}-b^{2})/2b$
for $r<b$. For our systems at room temperature the results were almost
indistinguishable for finite $b\le0.5$ so that we settled with $b=0.2$
in this work.

@@

The analytical expression for the excess chemical potential $\mu$
of an unstructured solute and the HNC closure has been derived by
Singer and Chandler (here in temperature units) \citep{Singer85}:
\begin{equation}
\mu_{\text{HNC}}=\rho\int d^{3}r\sum_{\alpha}\left[h_{\alpha}^{2}/2-c_{\alpha}-c_{\alpha}h_{\alpha}/2\right].\label{eq:chem-pot-hnc}
\end{equation}
A modification for the KH closure \citep{Kovalenko99} and a generalization
for the arbitrary closure of the form $h=f(x)$ with $x=-v+t$ are
available: 
\begin{equation}
\mu=\mu_{\text{HNC}}+\rho\int\sum_{\alpha}\phi(x_{\alpha})d^{3}r
\end{equation}
with $\phi(x)=f(x)-x-\int_{0}^{x}f(y)dy$ having particularly a simple
form for the PSE$n$ closures: $\phi_{n}(x)=-x^{n+1}/(n+1)!$ for
$x>0$ and zero otherwise \citep{Kast08}. In the case of the KH (PSE1)
closure the additional term $-x^{2}/2$ cancels the $h^{2}/2$ term
of the HNC integrand, Eq.~\ref{eq:chem-pot-hnc}, in the regions
where $x=h>0$.

@@

The free energy surface as a function of solute parameters $s$ is
constructed as a composite functional
\begin{equation}
G[n_{\text{e}},s]=E[n_{\text{e}},s]+\mu(s),
\end{equation}
where $E[n_{\text{e}},s]$ is the solute internal energy depending
on the solutes' geometric degrees of freedom $s$ and, in the case
of a QM solute, also on the electronic density $n_{\text{e}}$, but
not on the solvent site distribution functions $h$ \citep{Bo15}.

The uranyl solvation energy is then estimated as\textcolor{green}{{}
}
\begin{eqnarray}
\Delta G_{\text{sol}} & = & G[\text{UO}_{2}(\text{H}_{2}\text{O})_{n}^{2+}]-nG[\text{H}_{2}\text{O}]-E[\text{UO}_{2}^{2+}]\label{eq:dG-sol-definition}
\end{eqnarray}
for $n\ge0$ explicit aqua ligands. Note that without proper statistical
sampling over the solute degrees of freedom $s$ this approximation
of the solvation energy, justified to some extent for sufficiently
rigid solutes, breaks for increasingly many explicit water molecules
outside of the tightly bound first solvation shell as $n\to\infty$.
Indeed, in a gedankenexperiment the difference $G[(\text{H}_{2}\text{O})_{n}]-nG[\text{H}_{2}\text{O}]$
for the above definition of $G$ applied to water droplets asymptotically
approaches $n(\bar{E}[\text{H}_{2}\text{O}]-G[\text{H}_{2}\text{O}])$
where $\bar{E}$ is the average configurational energy of a water
molecule which is about $-$9.9~kcal/mol \citep{Berendsen87,Mahoney00}
to be compared to the solvation energy of $-$6.3~kcal/mol derived
from the ratio of the equilibrium vapor/liquid concentrations \citep{Bridgeman64}.

\subsection{Numerical solution of 3D RISM equations}

Because of the highly non-linear closure relation, a single variable
fix point iteration scheme, $t^{(n+1)}=T[t^{(n)}]$, with $T[t]=(\chi-1)\ast c(t)$
and $c(t)=f(-v+t)-t$ as derived from the non-linear equation system,
Eqs.~\ref{eq:rsim-h-c-t}, is bound to diverge for all but the simplest
cases. To overcome the divergence problem we solve the reduced non-linear
problem $F[t]=T[t]-t=0$ using the Newton--Krylov non-linear solver
\citep{petsc-user-ref} with the analytic matrix-free Jacobian, $J[t]=\delta F/\delta t$,
occasionally preceded by a few Picard iterations.

The long-range field of a charged solute requires a special treatment
in the 3D case. All of the practical approaches, including the Ng
scheme \citep{Ng74}, separate the smooth asymptote $v_{\text{L}}$
of the site specific potential off the indirect correlation $t$ as
well as $-v_{\text{L}}$ off the direct correlation $c$ to operate
primarily with the short range counterparts $v_{\text{S}}$ , $c_{\text{S}}$,
and $t_{\text{S}}$. An iteration scheme for the new primary variable
$t_{\text{S}}$ (and the resulting non-linear problem) will look slightly
different, though:

\begin{eqnarray}
T_{\text{S}}[t_{\text{S}}] & = & (\chi-1)\ast c_{\text{S}}(t_{\text{S}})-\tau
\end{eqnarray}
with $c_{\text{S}}(t_{\text{S}})=f(-v_{\text{S}}+t_{\text{S}})-t_{S}$
and the only term that involves a long-range operand being the constant
``renormalization'' term $\tau=\chi\ast v_{\text{L}}$. This term
is the convolution of solvent susceptibility with the asymptotic potential.

>>>>

%% LyX 2.0.6 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,journal=jpcafh,layout=twocolumn,maxauthors=0]{achemso}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{esint}
\usepackage[numbers]{natbib}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.

\title{Uranyl solvation by a three-dimensional reference interaction site
model}
\author{Alexei Matveev}
\email{alexei.matveev@gmail.com}
\author{Bo Li}
\author{Notker Rösch}
\email{roesch@mytum.de}
\affiliation{Department Chemie, Technische Universität München, 85747 Garching,
Germany}
\alsoaffiliation{Catalysis Research Center, Technische Universität München, 85747
Garching, Germany}
\alsoaffiliation{Institute of High Performance Computing, Agency for Science, Technology
and Research, 1 Fusionopolis Way, Connexis \#16-16, Singapore 138632,
Singapore}
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{mathptmx}
\usepackage{times}
%\usepackage{microtype}
%\def\thetable{\arabic{table}}
%\usepackage{ragged2e}
%\usepackage[textfont=it]{caption}
\captionsetup{font={bf}}
\usepackage{natmove}

@@

\usepackage{babel}
\begin{document}
\begin{abstract}
We report an implementation of the three-dimensional reference interaction
site model (3D RISM) that in particular addresses the treatment of
the long-range Coulomb field of charged species, represented by point
charges and/or a distributed charge density. A comparison of 1D and
3D results for atomic ions demonstrates a reasonable accuracy, even
for a moderate size of the unit cell and a moderate grid resolution.
In an application to uranyl complexes with 4\textendash{}6 explicit
aqua ligands and an implicit bulk solvent modeled by RISM, we show
that the 3D technique is not susceptible to the deficiencies of the
1D technique exposed in our previous work {[}Li, Matveev, Krüger,
Rösch, \emph{Comp. Theor. Chem.} \textbf{1051}, p.~151, (2015){]}.
The 3D method eliminates the artificial superposition of explicit
aqua ligands and the RISM medium and predicts essentially the same
values for uranyl and uranyl-water bond lengths as a state-of-the-art
polarizable continuum model. With the first solvation shell treated
explicitly, the observables are nearly independent of the order of
the closure relationship used when solving the set of integral equations
for the various distribution functions. Furthermore, we calculated
the activation barrier of water exchange with a hybrid approach that
combines the 3D RISM model for the bulk aqueous solvent and a quantum
mechanical description (at the level of electronic Density Functional
Theory) of uranyl interacting with explicitly represented water molecules.
The calculated result agrees very well with experiment and the best
theoretical estimates. 
\end{abstract}
\maketitle

@@

Thermodynamic and structural properties of the uranyl(VI) cation,
UO$_{2}^{2+}$, the most common uranium species in aqueous solution,
are crucial for understanding its environmental chemistry as well
as chemical processes in radioactive waste disposal and recycling
of used nuclear fuel \citep{book:Morss06,book:Grenthe92,Silva95}.
Simulations on the basis of quantum mechanics (QM) or classical molecular
mechanics (MM) provide two complimentary ways of addressing the microscopic
structure and the energetics of solvated uranyl species. Till now,
the most popular strategy for combining the accuracy of QM methods
and the complexity of the statistical solvent description is to apply
a polarizable continuum model (PCM) to represent the effective electrostatic
field of the solvent, usually employed in combination with phenomenological
corrections (largely relying on empirical parameters) for dispersion-repulsion
interactions and the cavity formation energy \citep{Tomasi05,Fuchs02}.
The conductor-like screening model (COSMO) of electrostatic interactions
is a state-of-the-art PCM \citep{Klamt95}; it was successfully applied
in various studies of uranyl complexes \citep{Fuchs02,Moskaleva04,Cao05,Shamov05,Gutowski06}.
On the other hand, significant efforts are being made on improving
the empirical model potentials for MM simulations, be it molecular
dynamics (MD) or Monte-Carlo (MC) methods, that yield the most explicit
statistical description of the solvent \citep{Guilbaud93,Guilbaud96,Rai12,Pomogaev13,Kerisit13,Tiwari14,book:Griebel07}.
Although several studies \citep{Gordon00,hagberg05,Frick09,Tirler13}
attempted to combine QM and stochastic MM models, such combinations
will inevitably lack the precise reproducibility and the efficiency
of analytic continuum models due to statistical sampling techniques.
More demanding QM dynamic simulations \citep{Buehl05,Buehl06,Buehl11,Nichols08}
are restricted to comparably small numbers of solvent molecules.

Alternatively, the integral equation theories of molecular liquids
\citep{book:Hirata03,book:Hansen06} may provide sufficient accuracy
at an affordable cost for simulating chemical processes in solution
by employing standard numerical methods on contemporary hardware.
Two variants of this type are the reference interaction site model
(RISM) \citep{Chandler72,Hirata81,Hirata82,Kinoshita97} and its three-dimensional
generalization (3D RISM) \citep{Beglov95,Kovalenko98}. These methods
provide the solvent structure in form of site-site radial distribution
functions (RDFs) or three-dimensional distribution functions of solvent
sites in an external (solute) field, respectively. Both 1D and 3D
methods offer variational expressions for the solvation energy in
the form of the excess chemical potential for several types of closure
of the hierarchy of equations \citep{Singer85,Kast08}, hence admit
intuitive expressions for the first-order derivatives with respect
to nuclear displacements. For another class of integral equations
based on the Born--Green--Yvon hierarchy \citep{Fischer80,Griebel08}
the lack of a practical free energy expression complicates exploring
free energy landscapes. Note that the dielectrically consistent RISM
(DRISM) \citep{Perkyns92} is able to compensate the error of the
gas-like dielectric response of standard RISM for electrolytes and
ionic solutes \citep{Perkyns94,Chiodo07,Chuev08,Joung13}.

Handling correctly the electrostatic field of a charge distribution
requires additional care when periodic boundary conditions are employed
to simulate finite systems \citep{Makov95}. It was understood early
on that the long-range asymptotes of the radial correlation functions
induced by Coulomb interactions in RISM need a separate treatment
as well \citep{Ng74}. The few practical approaches to 3D RISM pursue
various strategies, e.g.\ inclusion of a background charge correction
\citep{Kovalenko00b}, monopole correction \citep{Kloss08b}, and
an exact re-summation for a superposition of Gaussian-type charge
distributions \citep{Perkyns10}. Very recently, a new efficient technique
has been presented to avoid integrating in real space the long-range
contributions to the solvation energy \citep{Heil15}. In this work
we present a generalization of the exact re-summation technique for
arbitrary solute charge densities.

The integral equation theory has been combined with Hartree--Fock
electronic structure theory for self-consistent field calculations
(RISM SCF) in the pioneering work of Ten-no, Hirata, and Kato who
first applied the method to formaldehyde and carbonyl compounds in
water \citep{Tenno93,Tenno94}. An extension to 3D RISM SCF was developed
for density functional theory (DFT) \citep{Kovalenko99} and \emph{ab
initio} calculations \citep{Kloss08}. Analytic energy gradients with
respect to nuclear displacements for the self-consistent combination
of DFT and 3D RISM \citep{Gusarov06} simplify exploring free energy
landscapes of complex molecular systems. Some of the recent developments
of hybrid approaches combining integral equation theories and quantum
chemistry have been discussed in a recent review \citep{Sato13}.

In this work, we first describe our implementation of the 3D RISM
approach as well as a numerical treatment of the long-range Coulomb
field of a charged solute, Section~\ref{sec:methodology}. Then we
validate the implementation by comparing the results of 1D and 3D
techniques for alkali and halide ions, Section~\ref{sec:simple-ions}.
We further evaluate the performance of the method for the uranyl(VI)
dication in aqueous solution, represented by aqua complexes $[\text{UO}_{2}(\text{H}_{2}\text{O})_{n}]^{2+}$,
($n=4-6$), using a force field description, Section~\ref{sub:MM+RISM},
and a quantum mechanical representation of the interactions within
the first solvation shell, Section~\ref{sub:QM+RISM}. We also compare
the results to those of PCM calculations. Finally, we analyze the
activation barrier for exchanging a water molecule in the first solvation
shell, Section~\ref{sub:water-exchange}, using QM and MM models
of interactions. With the exception of the water exchange process
the test cases used to evaluate the 3D RISM approach are the same
as in our preceding work on 1D RISM \citep{Bo15}.

@@

In the 3D RISM approach, the solvent sites are exposed to an external
field of a solute of general shape. The solute-solvent interactions
are assumed to be pairwise, which in the present formulation excludes
\emph{polarizable} solutes.  Furthermore, the interactions are usually
approximated by superposition of pairwise site-site contributions.
The 3D RISM equations operating with site-specific distributions can
be written as follows \citep{book:Hirata03}:

\begin{equation}
\begin{cases}
h=\chi\ast c\\
h=c+t\\
h=f(-v+t)
\end{cases}\label{eq:rsim-h-c-t}
\end{equation}
Here $h$, $c$, and $t$ are the vectors of total, direct, and indirect
correlation functions $h_{\alpha}(\vec{r})$, $c_{\alpha}(\vec{r})$,
and $t_{\alpha}(\vec{r})$. All are dimensionless unknowns of the
equations that will be represented on a uniform grid of points. Greek
subscripts, e.g.\,$\alpha$, enumerate solvent interaction sites.

The solvent susceptibility $\chi$ in the first of the three Eqs.~\ref{eq:rsim-h-c-t}
is represented by a (symmetric square) matrix $\chi_{\alpha\gamma}(r)$
derived from the radial site-site correlation functions of the pure
solvent in 1D RISM equations for a specific temperature and solvent
number density $\rho$ \citep{book:Hirata03}. The star denotes the
convolution integral and a summation over corresponding matrix indices.
It is convenient to represent the solvent susceptibility by its (dimensionless)
Fourier image $\tilde{\chi}_{\alpha\gamma}(k)$ \citep{Chandler72,book:Hirata03}.
The solvent susceptibility as a function of either $r$ or $k$ is
spherically symmetric by construction.

The second of the Eqs.~\ref{eq:rsim-h-c-t} introduces the indirect
correlation $t$ that is conveniently used in the closure relation
--- the third and last of Eqs.~\ref{eq:rsim-h-c-t}. Closure is a
local relation between the values of the site potential $v_{\alpha}(\vec{r})$
and, in this case, the values of the indirect correlation $t_{\alpha}(\vec{r})$
and the direct correlation $h_{\alpha}(\vec{r})$, at each point $\vec{r}$.
It is the only non-linear relation between the unknowns $h$, $c$
and $t$ in Eqs.~\ref{eq:rsim-h-c-t}. The site-specific potentials
$v$ are measured in units of temperature and thus are also dimensionless.
For the purposes of this work we have chosen a very specific, though
popular, form of the closure, which happens to be a relation between
$h$ and the sum $x=-v+t$ sometimes referred to as renormalized indirect
correlation \citep{Kast08}:
\begin{equation}
f(x)=\begin{cases}
\mathrm{e}^{x}-1 & x\le0\\
\sum_{k=1}^{p}x^{k}/k! & x>0
\end{cases}\label{eq:closure-pse}
\end{equation}
In the limit of $p\to\infty$ one recovers the hypernetted chain (HNC)
closure \citep{Leeuwen59}. Kovalenko and Hirata (KH) proposed another
popular closure $(p=1)$ for improving the numerical stability of
the equations \citep{Kovalenko99}. Higher-order closure relations,
with $p>1$, are referred to as partial series expansion closures
of order $p$ (PSE$p$). For consistency, in this work we will refer
to the KH closure as PSE1.

We used pair-wise interactions between solute and solvent sites $i$
and $\alpha$ to define the solute field: $v_{\alpha}(\vec{r})=\beta\sum_{i}u_{i\alpha}(|\vec{r}-\vec{r}_{i}|)$
where $\beta$ is the inverse temperature. Each potential $u_{i\alpha}$
is defined as the sum of the Coulomb interaction $u_{i\alpha}^{\text{C}}(r)=q_{i}q_{\alpha}a[f_{\text{CS}}(ar)+f_{\text{CL}}(ar)]$,
where we separated the singular short-range part $f_{\text{CS}}(r)=\mathrm{erfc}(r)/r$
from the regular long-range part $f_{\text{CL}}(r)=\mathrm{erf}(r)/r$,
and a Lennard--Jones (LJ) type potential, $u_{i\alpha}^{\text{LJ}}(r)=\epsilon_{i\alpha}f_{\text{LJ}}(r/\sigma_{i\alpha})$
with $f_{\text{LJ}}(r)=4(r^{-12}-r^{-6})$. Parameters of the pair
interactions were derived from site parameters using the Lorentz--Berthelot
rules \citep{book:Hirata03}. The Ewald range parameter $a=1.2$~Å$^{-1}$
is common to all site pairs. To avoid numerical problems due to singularities,
the primitive forms $f_{\text{CS}}(r)$ and $f_{\text{LJ}}(r)$ were
regularized by replacing them by quadratic functions $\bar{f}(r)=f(b)+f^{\prime}(b)(r^{2}-b^{2})/2b$
for $r<b$. For our systems at room temperature the results were almost
indistinguishable for $0<b\le0.5$ so that we chose $b=0.2$.

@@

Singer and Chandler derived an analytical expression for the excess
chemical potential $\mu$ of an unstructured solute in the case of
HNC closure (in temperature units) \citep{Singer85}:
\begin{equation}
\mu_{\text{HNC}}=\rho\int d^{3}r\sum_{\alpha}\left[h_{\alpha}^{2}/2-c_{\alpha}-c_{\alpha}h_{\alpha}/2\right].\label{eq:chem-pot-hnc}
\end{equation}
A modification for the KH closure \citep{Kovalenko99} and a generalization
to an arbitrary closure of the form $h=f(x)$ with $x=-v+t$ are available:
\begin{equation}
\mu=\mu_{\text{HNC}}+\rho\int\sum_{\alpha}\phi(x_{\alpha})d^{3}r
\end{equation}
For PSE$p$ closures, $\phi(x)=f(x)-x-\int_{0}^{x}f(y)dy$ has a particularly
simple form: $\phi_{p}(x)=-x^{p+1}/(p+1)!$ for $x>0$ and zero otherwise
\citep{Kast08}. In the case of the KH (PSE1) closure, the additional
term $-x^{2}/2$ cancels the $h^{2}/2$ term of the HNC integrand,
Eq.~\ref{eq:chem-pot-hnc}, in the regions where $x=h>0$.

@@

The free energy surface as a function of solute parameters $s$ is
constructed as a composite functional
\begin{equation}
G[n_{\text{e}},s]=E[n_{\text{e}},s]+\mu(s),
\end{equation}
where $E[n_{\text{e}},s]$ is the solute internal energy that depends
on the solute's geometric degrees of freedom $s$ and, in the case
of a QM solute, also on the electronic density $n_{\text{e}}$, but
not on the solvent site distribution functions $h$ \citep{Bo15}.

The solvation energy of uranyl is estimated as\textcolor{green}{{} }
\begin{equation}
\Delta G_{\text{sol}}=G[\text{UO}_{2}(\text{H}_{2}\text{O})_{n}^{2+}]-nG[\text{H}_{2}\text{O}]-E[\text{UO}_{2}^{2+}]\label{eq:dG-sol-definition}
\end{equation}
for $n\ge0$ explicit aqua ligands. Note that without proper statistical
sampling over the solute degrees of freedom $s$ this approximation
of the solvation energy, justified to some extent for sufficiently
rigid solutes, does not hold for systems with a large number $n\to\infty$
of explicit water molecules outside of the tightly bound first solvation
shell. Indeed, in a gedanken experiment the difference $G[(\text{H}_{2}\text{O})_{n}]-nG[\text{H}_{2}\text{O}]$
for the above definition of $G$ applied to water droplets asymptotically
approaches $n(\bar{E}[\text{H}_{2}\text{O}]-G[\text{H}_{2}\text{O}])$
where $\bar{E}$ is the average configurational energy of a water
molecule; its value is about $-$9.9~kcal/mol \citep{Berendsen87,Mahoney00}.
This result is to be compared to the solvation energy of $-$6.3~kcal/mol
derived from the ratio of the equilibrium vapor/liquid concentrations
\citep{Bridgeman64}.

\subsection{Numerical solution of the 3D RISM equations}

Because of the highly non-linear closure relation, a single-variable
fix-point iteration scheme, $t^{(n+1)}=T[t^{(n)}]$, with $T[t]=(\chi-1)\ast c(t)$
and $c(t)=f(-v+t)-t$ as derived from the non-linear equation system,
Eqs.~\ref{eq:rsim-h-c-t}, is bound to diverge for all but the simplest
cases. To overcome the divergence problem we solve the reduced non-linear
problem $F[t]=T[t]-t=0$ using a Newton--Krylov non-linear solver
\citep{petsc-user-ref} with the analytic matrix-free Jacobian, $J[t]=\delta F/\delta t$;
the procedure is occasionally preceded by a few Picard iterations.

The long-range field of a charged solute requires a special treatment
in the 3D case. All practical approaches, including the scheme of
Ng \citep{Ng74}, separate the smooth asymptote $v_{\text{L}}$ of
the site-specific potential from the indirect correlation $t$ as
well as $-v_{\text{L}}$ from the direct correlation $c$ to operate
primarily with the short range counterparts $v_{\text{S}}$ , $c_{\text{S}}$,
and $t_{\text{S}}$. An iteration scheme for the new primary variable
$t_{\text{S}}$ (and the resulting non-linear problem) will look slightly
different:

\begin{equation}
T_{\text{S}}[t_{\text{S}}]=(\chi-1)\ast c_{\text{S}}(t_{\text{S}})-\tau
\end{equation}
with $c_{\text{S}}(t_{\text{S}})=f(-v_{\text{S}}+t_{\text{S}})-t_{S}$.
The constant ``renormalization'' term $\tau=\chi\ast v_{\text{L}}$,
the convolution of the solvent susceptibility with the asymptotic
potential, is the only one that involves a long-range operand.
